---
title: "Analyzing UK Smoking Data using PCA"
author: "Sejal Kankriya"
date: '2023-02-18'
output:
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
---

# PCA on UK Smoking Data

In this project we will analyze UK Smoking Data (`smoking.R`):

**Description**

Survey data on smoking habits from the UK. The data set can be used for
analyzing the demographic characteristics of smokers and types of
tobacco consumed.

**Format**

A data frame with 1691 observations on the following 12 variables.

`gender` - Gender with levels Female and Male.

`age` - Age.

`marital_status` - Marital status with levels Divorced, Married,
Separated, Single and Widowed.

`highest_qualification` - Highest education level with levels A Levels,
Degree, GCSE/CSE, GCSE/O Level, Higher/Sub Degree, No Qualification,
ONC/BTEC and Other/Sub Degree

`nationality` - Nationality with levels British, English, Irish,
Scottish, Welsh, Other, Refused and Unknown.

`ethnicity` - Ethnicity with levels Asian, Black, Chinese, Mixed, White
and Refused Unknown.

`gross_income` - Gross income with levels Under 2,600, 2,600 to 5,200,
5,200 to 10,400, 10,400 to 15,600, 15,600 to 20,800, 20,800 to 28,600,
28,600 to 36,400, Above 36,400, Refused and Unknown.

`region` - Region with levels London, Midlands & East Anglia, Scotland,
South East, South West, The North and Wales

`smoke` - Smoking status with levels No and Yes

`amt_weekends` - Number of cigarettes smoked per day on weekends.

`amt_weekdays` - Number of cigarettes smoked per day on weekdays.

`type` - Type of cigarettes smoked with levels Packets, Hand-Rolled,
Both/Mainly Packets and Both/Mainly Hand-Rolled

Source National STEM Centre, Large Datasets from stats4schools,
<https://www.stem.org.uk/resources/elibrary/resource/28452/large-datasets-stats4schools>.

Obtained from <https://www.openintro.org/data/index.php?data=smoking>

## Read and Clean the Data

```{r setup, include=FALSE,warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)


library(tibble)
library(readr)
library(dplyr)
library(stats)

library(broom)
library(cowplot)

library(ggplot2)
library(ggbiplot)
library(fastDummies)
library(forcats)
library(plotly)
```

```{r}
# Load data
source("smoking.R")
```

Take a look into data

```{r}
# place holder
smoking
# smoking_data
```

There are many fields there so for this exercise lets only concentrate
on smoke, gender, age, marital_status, highest_qualification and
gross_income.

Create new data.frame with only these columns.

```{r}
# place holder
df <- subset(smoking,
  select = c(
    "smoke", "gender", "age", "marital_status",
    "highest_qualification", "gross_income"
  ))
```

Omit all incomplete records

```{r}
# place holder
smoking_data <- na.omit(df)
```

For PCA feature should be numeric. Some of fields are binary (`gender`
and `smoke`) and can easily be converted to numeric type (with one and
zero). Other fields like `marital_status` has more than two categories,
convert them to binary (i.e. is_married, is_devorced). Several features
in the data set are ordinal (`gross_income` and
`highest_qualification`), convert them to some king of sensible level
(note that levels in factors are not in order)

```{r}
# place holder
smoking_data <- smoking_data %>% 
  mutate(
    gender = as.numeric(gender == "Female"),
    smoke = as.numeric(smoke == "Yes"),
    is_married = ifelse(marital_status == "Married", 1, 0),
    is_divorced = ifelse(marital_status == "Divorced", 1, 0),
    is_widowed = ifelse(marital_status == "Widowed", 1, 0),
    is_single = ifelse(marital_status == "Single", 1, 0),
    is_seperated = ifelse(marital_status == "Separated", 1, 0),
    gross_income = as.integer(as.factor(gross_income)),
    highest_qualification = as.integer(as.factor(highest_qualification))
) %>%
select(-marital_status)
```

PCA on all columns except smoking status

```{r echo=TRUE}
# place holder
pca_fit <- prcomp(smoking_data[c(-1)], scale=T)
pca_fit
pca_fit$sdev
```

Make a scree plot

```{r echo=TRUE, fig.height=4, fig.width=8}
# place holder
plot(pca_fit, type="line", main="Scree Plot")
```

Biplot color points by smoking field

```{r fig.width=5, fig.height=5}
ggbiplot(pca_fit, scale=0, groups = as.factor(smoking_data$smoke))
```

Based on the above biplot, there are two groups - smokers and
non-smokers. This can be pointed out by the different colors of the
points of two groups. The PC1 seems to be related to unmarried and
highest qualification. Whereas, the PC2 appears to be associated with
age.

We can use first two to discriminate smoking. However, just by using two
PCs wouldn't provide optimal seperation.

Based on the loading vector we can name the PC. Let's say if the PC1 is
associated with income and qualification we can name it as social class.
and if associated with marital status age, then we can name it as life
stage.

For highest_qualification variable, it may be more appropriate to assign
numbers for each levels according to their order. for eg. 1 for no
qualification, 2 for high school, etc.

Following the suggestion above and redo PCA and biplot

```{r}
source("smoking.R")
df <- subset(smoking,
  select = c(
    "smoke", "gender", "age", "marital_status",
    "highest_qualification", "gross_income"
  ))
smoking_data <- na.omit(df)
smoking_data_redo <- smoking_data %>% 
  mutate(
    gender = as.numeric(gender == "Female"),
    smoke = as.numeric(smoke == "Yes"),
    is_married = ifelse(marital_status == "Married", 1, 0),
    is_divorced = ifelse(marital_status == "Divorced", 1, 0),
    is_widowed = ifelse(marital_status == "Widowed", 1, 0),
    is_single = ifelse(marital_status == "Single", 1, 0),
    is_separated = ifelse(marital_status == "Separated", 1, 0),
    gross_income = as.integer(as.factor(gross_income)),
    highest_qualification = case_when(
      highest_qualification == "No Qualification" ~ 0,
      highest_qualification == "GCSE/O Level" ~ 1,
      highest_qualification == "Other/Sub Degree" ~ 2, 
      highest_qualification == "Higher/Sub Degree" ~ 3,
      highest_qualification == "Degree" ~ 4,
      highest_qualification == "A Levels" ~ 5,
      highest_qualification == "GCSE/CSE" ~ 6,
      highest_qualification == "ONC/BTEC" ~ 7)
  ) %>%
select(-marital_status)
```

```{r}
pca_fit_redo <- prcomp(smoking_data_redo[c(-1)], scale=T)
pca_fit_redo
pca_fit_redo$sdev
```

```{r fig.width=5, fig.height=5}
ggbiplot(pca_fit_redo, scale=0, groups = as.factor(smoking_data_redo$smoke))
```
